---
title: "Extended analyses for Northwest Territories caribou range planning"
author:
- affiliation: FOR-CAST Research & Analytics
  name: Alex M. Chubaty (achubaty@for-cast.ca)
- affiliation: University of British Columbia
  name: Tati Micheletti (tati.micheletti@gmail.com)
#abstract: ""
date: "`r format(Sys.time(), '%B %d, %Y')`"
#keywords: ""
#thanks: ""
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: templates/svm-latex-ms.tex
header-includes:
  - \usepackage{doi}
  - \usepackage{fancyhdr}
  - \usepackage{graphicx}
  - \usepackage{hyperref}
  - \usepackage{upquote}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \fancyfoot[R]{\thepage}
  - \fancypagestyle{plain}{\pagestyle{fancy}}
  - \setlength\headheight{72pt}
  - \rhead{\includegraphics[width=4cm]{images/logo.png}}
  - \bibpunct{(}{)}{;}{a}{}{,}
bibliography: references.bib
biblio-style: abbrvnat
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
endnote: no
fontfamily: mathpazo
fontsize: 11pt
geometry: margin=1in
#spacing: double
---

```{r setup, include=FALSE}
if (file.exists(".Renviron")) readRenviron(".Renviron")

source("01-packages.R")
source("02-init.R")
source("03-paths.R")

do.call(setPaths, summaryPaths)

Require(c("fs", "knitr", "qs"))

data_dirs <- list(
  landscape_raw = defaultPaths$outputPath,
  landscape = summaryPaths$outputPath,
  caribou = file.path(summaryPaths$outputPath, "caribouPlots"),
  caribouRSF_raw = file.path(summaryPaths$outputPath, "caribouRSF_predictions"),
  caribouPopGrowth_raw = posthocPaths$outputPath
)
stopifnot(dir.exists(unlist(data_dirs)))

## knitr options
knitr::opts_chunk$set(
  echo = FALSE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  results = "hold"
)
```

## Overview

We simulated vegetation, wildfire, and caribou in the NT1 study area for the years 2011-2100 following the modelling approaches of @micheletti2021 and @stewart, developed for an initial pilot study in the Northwest Territories.
All models have been developed using the PERFICT approach [@mcintire2022], which enables them to be part of update cycles as needed going forward.

Here, we used the most recent CIMP6 climate projections for four climate change scenarios (CanESM5 under each of SSP 370 and 585, as well as CNRM-ESM2-1 under each of SSP 370 and 585).
We generated wildfire predictions such that historical fire data from 2011-2020 are used instead of simulating fire in those years.
We ran 10 replicates for each scenario (instead of 5) to better capture stochastic variation for each scenario.
Caribou results were summarized over 10-year periods from 2011 to 2091, across each of the new or updated caribou planning polygons (provided by GNWT) within NT1 study area:

  a. Range planning areas (includes Wek'eezhii, Gwich’in, Inuvialuit, Sahtu, Southern NWT);
  b. Herd study areas;
  c. Geographic-barriers-to-caribou-movement areas;

We also summarized vegetation and wildfire model projections for the NT1 study area and for each of these new reporting areas over the entire simulation timescale (2011-2100).
Additional details are provided below.

### Code availability

Model code for this project is available from <https://github.com/FOR-CAST/NT_caribou> (public repository).

### Model outputs

A subset of data inputs and model outputs are available from <https://drive.google.com/drive/u/0/folders/0AAAMKzVjPKt5Uk9PVA> (restricted access) and are described in further detail below.

#### Filetypes

Please note that many of these output files are quite large due to the extent of the study area, and one might face software limitations when converting 
these objects to _e.g._, an Excel-compatible format.

- `.tif` files can be opened using the `raster` package in `R` or imported to GIS software (_e.g._, ArcGIS, QGIS);
- `.qs` files can be opened in `R` using the `qs` package (`qs::qread()`) and these tables may be subsequently saved in an alternate format (_e.g._, `.csv` for use with Microsoft Excel).

\newpage

## Landscape simulations

Simulations followed the approach described in @micheletti2021, but used the latest versions of the climate-sensitive vegetation model (_LandR.CS_), together with a climate-sensitive wildfire model (_fireSense_).
Additionally, we generated wildfire predictions such that historical fire data from 2011-2020 are used instead of simulating fire in those years.

Each simulation run produced the following files, where `YYYY` corresponds to the simulation year:

```{r outputs-landscape}
data.frame(
  filename = c(
    "ANPPMap_YYYY_yearYYYY.tif",
    "burnMap_YYYY_yearYYYY.tif",
    "burnSummary_year2100.qs",
    "cohortData_YYYY_yearYYYY.qs",
    "fireSense_EscapePredicted_YYYY_yearYYYY.tif",
    "fireSense_IgnitionPredicted_YYYY_yearYYYY.tif",
    "fireSense_SpreadPredicted_YYYY_yearYYYY.tif",
    "mortalityMap_YYYY_yearYYYY.tif",
    "pixelGroupMap_YYYY_yearYYYY.tif",
    "rstCurrentBurn_YYYY_yearYYYY.tif",
    "simulatedBiomassMap_YYYY_yearYYYY.tif"
  ),
  description = c(
    "yearly aboveground net primary productivity (ANPP) maps",
    "yearly wildfire burn maps",
    "wildfire burn summary for year 2100, containing information on all simulated fires with total area burned for all years",
    "yearly vegetation (tree only) cohort data objects, from which cohort age, aboveground biomass, and species composition can be extracted using the accompanying `pixelGroupMap`",
    "wildfire escape model predictions for all years",
    "wildfire ignition model predictions for all years",
    "wildfire spread model predictions for all years",
    "yearly vegetation (tree) mortality maps",
    "yearly map of vegetation pixel groups",
    "yearly wildfire of simulated burned pixels for all years",
    "yearly simulated vegetation biomass maps"
  )
) %>%
  knitr::kable(col.names = c("Filename pattern", "Description"), booktabs = TRUE) %>%
  kableExtra::kable_styling(font_size = 10, full_width = TRUE, latex_options = c("striped"))
```

&nbsp;

All landscape simulation summary products can be found at <https://drive.google.com/drive/u/0/folders/1t6032ggUC__jzaJs5H39LW6iFfrqlK_T> (restricted access).

\newpage

## Boreal caribou simulations

For both models below (RSF and Population Growth), we simulated changes across 90 years of climate-driven landscape change [as described in @stewart].
We used the  `caribouMetrics` package (<https://github.com/LandSciTech/caribouMetrics>) which implements caribou resource selection functions, as well as caribou population and demographic models (which extend the original models proposed by Johnson et al. [-@johnson2020]).

&nbsp;

All caribou summary products can be found at <https://drive.google.com/drive/u/0/folders/1IpC_u5c7Mluvqdxzk3pue5NFaBHyY6A6> (restricted access).

### Boreal caribou simulation: population resource selection (RSF)

We forecast suitable habitat for Boreal woodland caribou (_Rangifer tarandus caribou_) across the union of the three sets of polygons containing Northwest Territories' caribou ranges.
We generated the following products, where `SHP` stands for the shapefile used for the maps (`RangePlanRegions`, `GNWT_NT1_range`, `Revised_Study_Areas`), `YYYY` stands for the year (2011, 2031, 2051, 2071 or 2091), `RR` stands for the replicate number (01 to 10), `CLIMMOD` stands for the climate model (CanESM5 or CNRM-ESM2-1) and `XXX` stands for the SSP (370 or 585): 

```{r outputs-caribouRSF}
data.frame(
  filename = c(
    "averageDifference_SHP_allScenarios.png",
    "averageDifference_SHP_NT1_BCR6_CLIMMOD_SSPXXX.png",
    "meanRSFperPolygon_SHP.qs",
    "meanRSFperPolySummary_SHP.qs",
    "NT1_BCR6_CLIMMOD_SSPXXX_runRR_relativeSelectioncaribouRSF_NT_YearYYYY.tif",
    "NT1_BCR6_CLIMMOD_SSPXXX_runRR_relativeSelectionUncertaincaribouRSF_NT_YearYYYY.tif",
    "sdDifference_SHP_allScenarios.png",
    "sdDifference_SHP_NT1_BCR6_CLIMMOD_SSPXXX.png"
  ),
  description = c(
    "map of the average difference between first and last year across all climate scenarios, per shapefile",
    "map of the average difference between first and last year across, per climate scenario, per shapefile",
    "summarized table with mean, standard deviation (SD) and confidence interval (CI) of binned RSF values for each polygon within the shapefile, both for all and across climate model, across replicates",
    "table of mean, standard deviation (SD) and confidence interval (CI) of binned RSF values for each polygon within the shapefile, climate model and replicate within the shapefile",
    "bi-decadal relative selection (RSF) forecast for each climate model and replicate",
    "bi-decadal relative selection (RSF) forecast uncertainty (in the form of standard error) for each climate model and replicate",
    "map of the SD difference first and last year across all climate scenarios, per shapefile",
    "map of the SD difference first and last year, per climate scenario, per shapefile"
  )
) %>%
  knitr::kable(col.names = c("Filename pattern", "Description"), booktabs = TRUE) %>%
  kableExtra::kable_styling(font_size = 10, full_width = TRUE, latex_options = c("striped"))
```

\newpage

### Boreal caribou simulations: population growth

We also simulated demographic parameters for all polygons within all three shapefiles.
We generated the following products, where `SHP` stands for the shapefile used for the maps (`RangePlanRegions`, `GNWT_NT1_range`, `Revised_Study_Areas`), `YYYY` stands for the year (2011, 2031, 2051, 2071 or 2091), `RR` stands for the replicate number (01 to 10), `CLIMMOD` stands for the climate model (CanESM5 or CNRM-ESM2-1) and `XXX` stands for the SSP (370 or 585): 

```{r outputs-caribouLambda}
data.frame(
  filename = c(
    "caribouLayers_yearYYYY_NT1_BCR6_CLIMMOD_SSPXXX_runRR.tif",
    "disturbances_Year2091_NT1_BCR6_CLIMMOD_SSPXXX_runRR_year2091.rds",
    "predictedCaribou_Year2091_NT1_BCR6_CLIMMOD_SSPXXX_runRR_year2091.rds",
    "caribou_SHP_allCM_Johnson_Johnson.png",
    "populationGrowthTable.csv",
    "disturbanceTable.csv"
  ),
  description = c(
    "layers used for population growth calculation per climate model and replicate",
    "disturbance summary at year 2091 (which contains all disturbances for all other years), per climate model and replicate",
    "predicted caribou values (lambda, recruitment and adult female mortality) at year 2091 (which contains values for all other years), per climate model and replicate",
    "caribou population demographics plots per polygon per shapefile, showing increasing levels of uncertainty (5 to 85 quantiles; coloured bands) using climate models and replicates as uncertainty measures",
    "caribou population demographics table (summary) per polygon per shapefile, replicate and climate model",
    "disturbance table (summary) per polygon per shapefile, replicate and climate model"
  )
) %>%
  knitr::kable(col.names = c("Filename pattern", "Description"), booktabs = TRUE) %>%
  kableExtra::kable_styling(font_size = 10, full_width = TRUE, latex_options = c("striped"))
```

<!-- References: autogenerated from references.bib -->
